---
title: "Budyko relationship"
author: "Francesco Grossi and Beni Stocker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P-model usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Aim

This vignette illustrate the Budyko relationship calculated based on interannual mean of precipitation, condensation, actual and potential evapotranspiration obtained from local measurment and simulated using P model.

This vignette highlits the difference in P model simulation by considering only the top 2-meter soil (referred as "costant") and by using the water holding capacity from [Stocker et al.](https://www.nature.com/articles/s41561-023-01125-2#ref-CR1) (referred as "variable")

```{r}
# library and data loading
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
library(cowplot)
library(FluxDataKit)
costant_whc <- readRDS(here("data","output_costant_whc.rds"))
variable_whc <- readRDS(here("data","output_variable_whc.rds"))
```


### Budyko relationship and mathematical fitting

The data are fitted using the Fu equation, it has one free parameter $\omega$ which represent all the physico and biological charachteristc of the catchment.

$$
AET/precipitation = 1 + PET/precipitation - (1 + (PET/precipitation)^\omega)^{1/\omega}
$$
The equation will be used to fit all the data point, in the report "Fu_equation" 
different subset of data will be used.

```{r}
fit_fun = function(x,omega){
  return( 1 + x - (1 + (x)^omega)^(1/omega))}
```

```{r}
# fit with old whc
old_fit <- nls(aet/prec_cond ~ (1 + pet/prec_cond)-(1 + (pet/prec_cond)^a)^(1/a)
               , data = costant_whc,start = list(a = 2.3))

plot_fit <- data.frame(obs = costant_whc$aet/costant_whc$prec_cond, pred= fit_fun(costant_whc$pet/costant_whc$prec_cond, coef(old_fit)))

param = lm(pred ~ obs, data= plot_fit)

r_squared <- as.numeric(format(cor(plot_fit$obs ,plot_fit$pred, 
                                              use = "complete.obs")^2, digits=2))
coeff <- as.numeric(format(coef(old_fit), digits=2))
slope <- as.numeric(format(param$coefficients[2],digits=2))
intercept <- as.numeric(format(param$coefficients[1],digits=2))


plot_1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  labs( title = "Fu equation fit with costant whc", 
        subtitle = bquote("R "^2*" = "~.(r_squared)~
                          " omega = "~.(coeff)~
                          " slope = "~.(slope)
                         # ~" int = "~.(intercept)
                          )) +
  xlab("observed") +
  ylab("predicted")

# fit with new whc
new_fit <- nls(aet/prec_cond ~ 
                 (1 + pet/prec_cond)-(1 + (pet/prec_cond)^a)^(1/a), 
               data = variable_whc,start = list(a = 2.3))

plot_fit <- data.frame(obs = variable_whc$aet/variable_whc$prec_cond, pred= fit_fun(variable_whc$pet/variable_whc$prec_cond, coef(new_fit)))

param = lm(pred ~ obs, data= plot_fit)

r_squared <- as.numeric(format(cor(plot_fit$obs ,plot_fit$pred, 
                                              use = "complete.obs")^2, digits=2))
coeff <- as.numeric(format(coef(new_fit), digits=2))
slope <- as.numeric(format(param$coefficients[2],digits=2))
intercept <- as.numeric(format(param$coefficients[1],digits=2))


plot_2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
    labs( title = "Fu equation fit with variable whc", 
        subtitle = bquote("R "^2*" = "~.(r_squared)~
                          " omega = "~.(coeff)~
                          " slope = "~.(slope)
                         # ~" int = "~.(intercept)
                          )) +
  ylab("predicted")+
  xlab("observed")

plot_grid(plot_1, plot_2)
```

```{r}
plot_1 <- ggplot(costant_whc) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(costant_whc)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(costant_whc)[1]), coef(old_fit)))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA) + 
  xlab("PET / prec") +
  ylab("AET / prec") + 
  ggtitle("Budyko curve with costant WHC")

plot_2 <- ggplot(variable_whc) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(variable_whc)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(variable_whc)[1]), coef(new_fit)))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA) + 
  xlab("PET / prec ") +
  ylab("AET / prec ") + 
  ggtitle("Budyko curve with variable WHC")

plot_grid(plot_1, plot_2)
```
