<<<<<<< HEAD
---
title: "Budyko curve"
output: html_document
date: "2024-02-28"
editor_options: 
  markdown: 
    wrap: sentence
---

## Aim

The aim of this vignette is to create the Budyko curve starting from the data provided on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081)).
To use the vignette properly, download the data and inserted in the folder [data](https://github.com/stineb/rsofundemo/tree/main/data).
The vignette will use the {rsofun} package and will run the p model as the vignette "p_model_multisite.Rmd".

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rsofun)
library(here)
```

```{r}
# use the data found on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081))
driver <- readRDS(here("data/rsofun_driver_data_clean.rds"))
```

```{r}
# run the model as in the previous vignette
params_modl <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)

# run the model for these parameters
output <- rsofun::runread_pmodel_f(
  driver,
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
output <- output |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len != 1) |>
  select(-len)
```

## Multi year averaging

The Budyko relationship is based on long term measurement.
In this vignette the data will be averaged over the entire time of data collection.

```{r}
# average pet and aet over years 
get_annual_aet_pet <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(aet = sum(aet),
              pet = sum(pet)) |>
    ungroup() |>
    summarise(aet = mean(aet),
              pet = mean(pet))
}

adf <- output |>
  mutate(adf = purrr::map(data, ~get_annual_aet_pet(.))) |>
  unnest(adf) |>
  select(sitename, aet, pet)

# add annual precipitation from forcing
get_annual_prec_cond <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(prec_cond = sum(prec_cond)) |>
    ungroup() |>
    summarise(prec_cond = mean(prec_cond))
}

adf <- driver |>
  unnest(forcing) |>
  left_join(
    output |>
      unnest(data) |>
      select(-snow, -netrad, -fapar, gpp_pmodel = gpp),
    by = c("sitename", "date")
  ) |>
  mutate(prec = (rain + snow) * 60 * 60 * 24) |>
  mutate(prec_cond = prec + cond) |>
  group_by(sitename) |>
  nest() |>
  mutate(adf = purrr::map(data, ~get_annual_prec_cond(.))) |>
  unnest(adf) |>
  select(sitename, prec_cond) |>
  right_join(
    adf,
    by = "sitename"
  )
```

## AET vs precipitation

This figure illustrate the relationship of the AET and precipitation.
Each point represent the multi-year average on a site of AET and precipitation.
Ideally, all the AET value shuold be equal or less than the precipitation value.

```{r}
# AET vs. precipitation plus condensation
adf |>
  ggplot(aes(prec_cond, aet)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  xlim(0, NA) +
  ylim(0, NA)
```

## Curve fitting

The Budyko curve can be modelled following the equation obtain by [FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03) (chinese only).

$$
AET = 1 + PET - (1 + (PET)^\omega)^{1/\omega}
$$

To fit the data, a non-linear least square method is used (refer to {nls}).
The only free parameter is $\omega$.
The fitted equation provides a good estimation of the data, with an r squared of 0.83.

```{r}
# curve fitting and adf updating
#formula: y = 1 + x -( 1+ (x)^omega)^1/omega 
# ([FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03)) chinese only 

fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = adf,start = list(a = 2.3))
fit_fun = function(x){
  return( 1 + x - (1 + (x)^coef(fit))^(1/coef(fit)))}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))


```

## Budyko curve

The Budyko curve is shown in this figure along with the fitted model

```{r}
# plotting results

ggplot(adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA)
```

## WHC fitting

Fitting taking in account the water holding capacity (whc) (eq 14) in [Mianabadia, 2020](https://www.sciencedirect.com/science/article/abs/pii/S0022169420305497).

$$
AET = (e^{\alpha*(1-PET^{-1})} -1 )/( e^{\alpha*(1-PET^{-1})} -  PET^{-1} )
$$

Where $\alpha$ is the ratio between the whc and the mean storm depth.
Two test will be performed

-   the first taking $\alpha$ as free parameter (implying whc uniform)
-   the second taking $\alpha$ as whc / free parameter

In both cases the mean storm depth is taken as free parameter.
The first case perform slightly better having a R\^2 of 0.76 compared to 0.72.

```{r}
fit = nls(aet/prec_cond ~ (exp(a*(1-(pet/prec_cond)^-1))-1)/(exp(a*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp(coef(fit)*(1-x^-1))-1)/(exp(coef(fit)*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))


```

```{r}
ggplot(adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0,4,length.out=dim(adf)[1]),
                y = fit_fun(seq(0,4,length.out=dim(adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA)
```

```{r}
whc <-  output |>
  unnest(site_info) |>
  select(whc)

adf$whc <- whc[[1]]

fit = nls(aet/prec_cond ~ (exp((whc/a)*(1-(pet/prec_cond)^-1))-1)/(exp((whc/a)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp((adf$whc/coef(fit))*(1-x^-1))-1)/(exp((adf$whc/coef(fit))*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))

```

```{r}
adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) 
```
=======
---
title: "Budyko curve"
output: html_document
date: "2024-02-28"
editor_options: 
  markdown: 
    wrap: sentence
---

## Aim

The aim of this vignette is to create the Budyko curve starting from the data provided on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081)).
To use the vignette properly, download the data and inserted in the folder [data](https://github.com/stineb/rsofundemo/tree/main/data).
The vignette will use the {rsofun} package and will run the p model as the vignette "p_model_multisite.Rmd".

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rsofun)
library(here)
```

```{r}
# use the data found on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081))
driver <- readRDS(here("data","rsofun_driver_data_clean.rds"))
```

```{r}
# run the model as in the previous vignette
params_modl <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)

# run the model for these parameters
output <- rsofun::runread_pmodel_f(
  driver,
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
output <- output |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len != 1) |>
  select(-len)
```

## Multi year averaging

The Budyko relationship is based on long term measurement.
In this vignette the data will be averaged over the entire time of data collection.

```{r}
# average pet and aet over years 
get_annual_aet_pet <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(aet = sum(aet),
              pet = sum(pet)) |>
    ungroup() |>
    summarise(aet = mean(aet),
              pet = mean(pet))
}

adf <- output |>
  mutate(adf = purrr::map(data, ~get_annual_aet_pet(.))) |>
  unnest(adf) |>
  select(sitename, aet, pet)

# add annual precipitation from forcing
get_annual_prec_cond <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(prec_cond = sum(prec_cond)) |>
    ungroup() |>
    summarise(prec_cond = mean(prec_cond))
}

adf <- driver |>
  unnest(forcing) |>
  left_join(
    output |>
      unnest(data) |>
      select(-snow, -netrad, -fapar, gpp_pmodel = gpp),
    by = c("sitename", "date")
  ) |>
  mutate(prec = (rain + snow) * 60 * 60 * 24) |>
  mutate(prec_cond = prec + cond) |>
  group_by(sitename) |>
  nest() |>
  mutate(adf = purrr::map(data, ~get_annual_prec_cond(.))) |>
  unnest(adf) |>
  select(sitename, prec_cond) |>
  right_join(
    adf,
    by = "sitename"
  )
```

## AET vs precipitation

This figure illustrate the relationship of the AET and precipitation.
Each point represent the multi-year average on a site of AET and precipitation.
Ideally, all the AET value shuold be equal or less than the precipitation value.

```{r}
# AET vs. precipitation plus condensation
adf |>
  ggplot(aes(prec_cond, aet)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  xlim(0, NA) +
  ylim(0, NA)
```

## Curve fitting

The Budyko curve can be modelled following the equation obtain by [FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03) (chinese only).

$$
AET = 1 + PET - (1 + (PET)^\omega)^{1/\omega}
$$

To fit the data, a non-linear least square method is used (refer to {nls}).
The only free parameter is $\omega$.
The fitted equation provides a good estimation of the data, with an r squared of 0.83.

```{r}
# curve fitting and adf updating
#formula: y = 1 + x -( 1+ (x)^omega)^1/omega 
# ([FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03)) chinese only 

fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = adf,start = list(a = 2.3))
fit_fun = function(x){
  return( 1 + x - (1 + (x)^coef(fit))^(1/coef(fit)))}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))


```

## Budyko curve

The Budyko curve is shown in this figure along with the fitted model

```{r}
# plotting results

ggplot(adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA)
```

## WHC fitting

Fitting taking in account the water holding capacity (whc) (eq 14) in [Mianabadia, 2020](https://www.sciencedirect.com/science/article/abs/pii/S0022169420305497).

$$
AET = (e^{\alpha*(1-PET^{-1})} -1 )/( e^{\alpha*(1-PET^{-1})} -  PET^{-1} )
$$

Where $\alpha$ is the ratio between the whc and the mean storm depth.
Two test will be performed

-   the first taking $\alpha$ as free parameter (implying whc uniform)
-   the second taking $\alpha$ as whc / free parameter

In both cases the mean storm depth is taken as free parameter.
The first case perform slightly better having a R\^2 of 0.76 compared to 0.72.

```{r}
fit = nls(aet/prec_cond ~ (exp(a*(1-(pet/prec_cond)^-1))-1)/(exp(a*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp(coef(fit)*(1-x^-1))-1)/(exp(coef(fit)*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))


```

```{r}
ggplot(adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0,4,length.out=dim(adf)[1]),
                y = fit_fun(seq(0,4,length.out=dim(adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA)
```

```{r}
whc <-  output |>
  unnest(site_info) |>
  select(whc)

adf$whc <- whc[[1]]

fit = nls(aet/prec_cond ~ (exp((whc/a)*(1-(pet/prec_cond)^-1))-1)/(exp((whc/a)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp((adf$whc/coef(fit))*(1-x^-1))-1)/(exp((adf$whc/coef(fit))*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))

```

```{r}
adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) 
```


By trying to parametrize the "mean storm depth" as prec_cond yields better result


```{r}

# adding prec_cond to the model

fit = nls(aet/prec_cond ~ (exp((whc/(prec_cond*a))*(1-(pet/prec_cond)^-1))-1)/(exp((whc/(prec_cond*a))*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp((adf$whc/(adf$prec_cond*coef(fit)))*(1-x^-1))-1)/(exp((adf$whc/(adf$prec_cond*coef(fit)))*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))

```



```{r}
adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) 
```

## New whc 

Test the whc obtained from [Stocker, 2023](https://zenodo.org/records/5515246)
Download the file "cwdx80.nc" and put in the data folder

```{r}
# updating whc
library(ncdf4)

nc_file <- nc_open(here("data","cwdx80.nc"))

whc = ncvar_get(nc_file, "cwdx80")
lons = ncvar_get(nc_file, "lon")
lats = ncvar_get(nc_file, "lat")

geo <- output |>
  unnest(site_info) |>
  select(lon  , lat)

geo$sitename <- output$sitename 

n <- 1 # parameter to select size of slice to average
new_whc <- lapply(geo$sitename, function(x){
  geo <- geo[geo$sitename == x,]
  lonid <- which(lons > geo$lon)[1]
  latid <- which(lats > geo$lat)[1]
  zroot_region <- whc[(lonid-n):(lonid+n), (latid-n):(latid+n)]
  whc_site <- mean(as.numeric(zroot_region, na.rm=T))
  return(whc_site)
  })
adf$whc <- unlist(new_whc)
```

```{r}
# plot to highlights the difference between old and new whc
old_whc <- output |>
  unnest(site_info) |>
  select(whc)
old_whc <- old_whc[[1]]
ggplot(data.frame(old = old_whc, new =  unlist(new_whc)), aes(x=old,y=new))+
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted")
```

```{r}
fit = nls(aet/prec_cond ~ (exp((whc/a)*(1-(pet/prec_cond)^-1))-1)/(exp((whc/a)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp((adf$whc/coef(fit))*(1-x^-1))-1)/(exp((adf$whc/coef(fit))*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))

```

```{r}
adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) 
```
>>>>>>> 89540b53710f1b0acfb1dfffd6dca69a0708a3d6
