# Budyko comparison

Fit comparison between older whc and newer whc.
This vignette has the only purpose to show the difference between the new and old whc dataset.
To have an in depth look at the models, refer to the others vignettes (ongoing).
All the file are generated by the script create_file_for_vignette.R

```{r message=FALSE}
# library loading
library(dplyr)
library(tidyr)
library(ggplot2)
library(here)
# to schow multiple plot
if (!require('gridExtra')) install.packages('gridExtra') 
library(gridExtra)
```

```{r}
# file loading for the vignette
old_adf <- readRDS(here("data","budyko_old_whc.rds"))
new_adf <- readRDS(here("data","budyko_new_whc.rds"))
veg <- readRDS(here("data","vegetation.rds"))
```

## AET and PET comparison

```{r warning=FALSE}
# AET new vs. old 
data.frame(old_AET = old_adf$aet/old_adf$prec_cond, 
           new_AET = new_adf$aet/new_adf$prec_cond) |>
  ggplot(aes(old_AET, new_AET)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  xlim(0, NA) +
  ylim(0, NA)

```
The AET was overestimated.

## Curve fitting

both dataframes will be fitted and plotted side by side

### Model 1

[FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03) (chinese only).

$$
PET = pet/precipitation\\
AET = 1 + PET - (1 + (PET)^\omega)^{-\omega}
$$

```{r}
fit_fun = function(x){
  return( 1 + x - (1 + (x)^coef(fit))^(1/coef(fit)))}
```

```{r}
# fit with old whc
fit <- nls(aet/prec_cond ~ (exp(a*(1-(pet/prec_cond)^-1))-1)/(exp(a*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)), data = old_adf,start = list(a = 2.3))

plot_fit <- data.frame(obs = old_adf$aet/old_adf$prec_cond, pred= fit_fun(old_adf$pet/old_adf$prec_cond))

plot1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  geom_text(x= 0.65, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.65, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("Old whc")
```

```{r}
# fit with new whc
fit <- nls(aet/prec_cond ~ (exp(a*(1-(pet/prec_cond)^-1))-1)/(exp(a*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)), data = new_adf,start = list(a = 2.3))

plot_fit <- data.frame(obs = new_adf$aet/new_adf$prec_cond, pred= fit_fun(new_adf$pet/new_adf$prec_cond))

plot2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.6, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("New whc")
```

```{r message=FALSE, warning=FALSE}
grid.arrange(plot1, plot2, ncol=2)
```

```{r}
plot1 <- ggplot(old_adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(old_adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(old_adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("Old whc")
```


```{r message=FALSE, warning=FALSE}
plot2 <- ggplot(new_adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(new_adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(new_adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("New whc")
grid.arrange(plot1, plot2, ncol=2)
```

## Fit with whc in the model

(eq 14) in [Mianabadia, 2020](https://www.sciencedirect.com/science/article/abs/pii/S0022169420305497).

$$
WHC = whc/precipitation\\
AET = (e^{\alpha*WHC*(1-PET^{-1})} -1 )/( e^{\alpha*WHC*(1-PET^{-1})} -  PET^{-1} )
$$

```{r message=FALSE, warning=FALSE}
fit = nls(aet/prec_cond ~ (exp((a*whc/(prec_cond)*(1-(pet/prec_cond)^-1)))-1)/(exp(a*whc/(prec_cond)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = old_adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = old_adf$aet/old_adf$prec_cond, pred= fit_fun(old_adf$pet/old_adf$prec_cond))

plot1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.7, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.65, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("Old whc")
```

```{r message=FALSE, warning=FALSE}
fit = nls(aet/prec_cond ~ (exp((a*whc/(prec_cond)*(1-(pet/prec_cond)^-1)))-1)/(exp(a*whc/(prec_cond)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = new_adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = new_adf$aet/new_adf$prec_cond, pred= fit_fun(new_adf$pet/new_adf$prec_cond))

plot2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.65, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("New whc")
```

```{r message=FALSE, warning=FALSE}
grid.arrange(plot1, plot2, ncol=2)
```

```{r message=FALSE, warning=FALSE}
fit_fun = function(x){
  (exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-x^-1)}

old_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("Old whc")
```

```{r message=FALSE, warning=FALSE}
fit_fun = function(x){
  (exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-x^-1)}

new_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA)  +
  ggtitle("New whc")
```
## Model with vegetation

Starting from the Fu equation, the alpha parameter varying ranging from 2.4 for savannas up to 4 for croplands in new whc and from 3.7 to 9 in old whc.
I create some small data frames in which each type is represented at least 5 times and merge again together.
Each data frames is fitted with his own coefficient.

```{r}
new_adf <- left_join(new_adf,veg,by="sitename")
old_adf <- left_join(old_adf,veg,by="sitename")
# vegetation filter (is the same for both)

type <- unique(old_adf[!is.na(old_adf$veg),]$veg)
count <- rep(0,length(type))
for(i in 1:216){
  if(!is.na(old_adf$veg[i])){
      count <- count + (type == old_adf$veg[i])}}

```

```{r}
# old whc filtering

old_adf <- old_adf[old_adf$veg %in% type[count>5],]

old_param <- NULL
for(i in type[count>5]){
  df <- old_adf[old_adf$veg == i,]
  fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = df,start = list(a = 2.5))
    tmp <- data.frame(type = i, coef = coef(fit))
  old_param <- rbind(old_param,tmp)
}

coeff <-c()
for(i in old_adf$veg){
  tmp <- old_param[old_param$type==i,2]
  coeff <- append(coeff,tmp)
}
old_adf$coef <- coeff

fit_fun = function(x,y){
  return( 1 + x - (1 + (x)^y)^(1/y))}

plot_fit = data.frame(obs = old_adf$aet/old_adf$prec_cond, 
                      pred= fit_fun(old_adf$pet/old_adf$prec_cond,old_adf$coef),
                      type= old_adf$veg)

plot1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.65, y=0.9,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4))) +
  ggtitle("Old whc")
```

```{r}
# new whc filtering

new_adf <- new_adf[new_adf$veg %in% type[count>5],]

new_param <- NULL
for(i in type[count>5]){
  df <- new_adf[new_adf$veg == i,]
  fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = df,start = list(a = 2.5))
    tmp <- data.frame(type = i, coef = coef(fit))
  new_param <- rbind(new_param,tmp)
}

# coefficent updating
coeff <-c()
for(i in new_adf$veg){
  tmp <- new_param[new_param$type==i,2]
  coeff <- append(coeff,tmp)
}
new_adf$coef <- coeff


plot_fit = data.frame(obs = new_adf$aet/new_adf$prec_cond, 
                      pred= fit_fun(new_adf$pet/new_adf$prec_cond,new_adf$coef),
                      type= new_adf$veg)

plot2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=0.9,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  ggtitle("New whc")
```

```{r message=FALSE, warning=FALSE}
grid.arrange(plot1, plot2, ncol=2)
```

```{r message=FALSE, warning=FALSE}
old_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond,coef)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("Old whc")
```

```{r message=FALSE, warning=FALSE}
new_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond,coef)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("New whc")

```

