# Budyko comparison

Fit comparison between older whc and newer whc.
This vignette has the only purpose to show the difference between the new and old whc dataset.
To have an in depth look at the models, refer to the others vignettes (ongoing)

```{r message=FALSE}
# library loading
library(dplyr)
library(tidyr)
library(ggplot2)
library(rsofun)
library(here)
library(ncdf4)
# to schow multiple plot
if (!require('gridExtra')) install.packages('gridExtra') 
library(gridExtra)
```

```{r}
# funuction used to create dataframes
get_annual_aet_pet <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(aet = sum(aet),
              pet = sum(pet)) |>
    ungroup() |>
    summarise(aet = mean(aet),
              pet = mean(pet))
}

get_annual_prec_cond <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(prec_cond = sum(prec_cond)) |>
    ungroup() |>
    summarise(prec_cond = mean(prec_cond))
}
```


```{r}
# paramter for p model
params_modl <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)
```

```{r warning=FALSE}
# run p model
old_driver <- readRDS(here("data","rsofun_driver_data_clean.rds"))

old_output <- rsofun::runread_pmodel_f(
  old_driver,
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
old_output <- old_output |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len != 1) |>
  select(-len)
```

```{r}
# create dataframe
old_adf <- old_output |>
  mutate(old_adf = purrr::map(data, ~get_annual_aet_pet(.))) |>
  unnest(old_adf) |>
  select(sitename, aet, pet)


old_adf <- old_driver |>
  unnest(forcing) |>
  left_join(
    old_output |>
      unnest(data) |>
      select(-snow, -netrad, -fapar, gpp_pmodel = gpp),
    by = c("sitename", "date")
  ) |>
  mutate(prec = (rain + snow) * 60 * 60 * 24) |>
  mutate(prec_cond = prec + cond) |>
  group_by(sitename) |>
  nest() |>
  mutate(old_adf = purrr::map(data, ~get_annual_prec_cond(.))) |>
  unnest(old_adf) |>
  select(sitename, prec_cond) |>
  right_join(
    old_adf,
    by = "sitename"
  )

whc <-  old_output |>
  unnest(site_info) |>
  select(whc)

old_adf$whc <- whc[[1]]
```

## Update whc

```{r}
# updating whc

nc_file <- nc_open(here("data","cwdx80.nc"))

whc = ncvar_get(nc_file, "cwdx80")
lons = ncvar_get(nc_file, "lon")
lats = ncvar_get(nc_file, "lat")

geo <- old_driver |>
  unnest(site_info) |>
  select(lon  , lat)

geo$sitename <- old_driver$sitename 

n <- 1 # parameter to select size of slice to average

new_whc <- lapply(geo$sitename, function(x){
  geo <- geo[geo$sitename == x,]
  lonid <- which(lons > geo$lon)[1]
  latid <- which(lats > geo$lat)[1]
  zroot_region <- whc[(lonid-n):(lonid+n), (latid-n):(latid+n)]
  whc_site <- mean(as.numeric(zroot_region, na.rm=T))
  return(whc_site)
  })

new_whc <- unlist(new_whc)

new_driver <- old_driver

for(i in 1:241){
  new_driver$site_info[i][[1]][4] <- new_whc[i]
}
rm(whc) # for memnory
```

```{r warning=FALSE}
# run p model
new_output <- rsofun::runread_pmodel_f(
  new_driver,
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
new_output <- new_output |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len != 1) |>
  select(-len)
```

```{r}
# create dataframe
new_adf <- new_output |>
  mutate(new_adf = purrr::map(data, ~get_annual_aet_pet(.))) |>
  unnest(new_adf) |>
  select(sitename, aet, pet)


new_adf <- new_driver |>
  unnest(forcing) |>
  left_join(
    new_output |>
      unnest(data) |>
      select(-snow, -netrad, -fapar, gpp_pmodel = gpp),
    by = c("sitename", "date")
  ) |>
  mutate(prec = (rain + snow) * 60 * 60 * 24) |>
  mutate(prec_cond = prec + cond) |>
  group_by(sitename) |>
  nest() |>
  mutate(new_adf = purrr::map(data, ~get_annual_prec_cond(.))) |>
  unnest(new_adf) |>
  select(sitename, prec_cond) |>
  right_join(
    new_adf,
    by = "sitename"
  )

whc <-  new_output |>
  unnest(site_info) |>
  select(whc)

new_adf$whc <- whc[[1]]
```

## AET and PET comparison

```{r warning=FALSE}
# AET new vs. old 
data.frame(old_AET = old_adf$aet/old_adf$prec_cond, 
           new_AET = new_adf$aet/new_adf$prec_cond) |>
  ggplot(aes(old_AET, new_AET)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  xlim(0, NA) +
  ylim(0, NA)

```
The AET was overestimated.

## Curve fitting

both dataframes will be fitted and plotted side by side

### Model 1

[FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03) (chinese only).

$$
PET = pet/precipitation\\
AET = 1 + PET - (1 + (PET)^\omega)^{-\omega}
$$

```{r}
fit_fun = function(x){
  return( 1 + x - (1 + (x)^coef(fit))^(1/coef(fit)))}
```

```{r}
# fit with old whc
fit <- nls(aet/prec_cond ~ (exp(a*(1-(pet/prec_cond)^-1))-1)/(exp(a*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)), data = old_adf,start = list(a = 2.3))

plot_fit <- data.frame(obs = old_adf$aet/old_adf$prec_cond, pred= fit_fun(old_adf$pet/old_adf$prec_cond))

plot1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  geom_text(x= 0.65, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.65, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("Old whc")
```

```{r}
# fit with new whc
fit <- nls(aet/prec_cond ~ (exp(a*(1-(pet/prec_cond)^-1))-1)/(exp(a*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)), data = new_adf,start = list(a = 2.3))

plot_fit <- data.frame(obs = new_adf$aet/new_adf$prec_cond, pred= fit_fun(new_adf$pet/new_adf$prec_cond))

plot2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE) + 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.6, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("New whc")
```

```{r message=FALSE, warning=FALSE}
grid.arrange(plot1, plot2, ncol=2)
```

```{r}
plot1 <- ggplot(old_adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(old_adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(old_adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("Old whc")
```


```{r message=FALSE, warning=FALSE}
plot2 <- ggplot(new_adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(new_adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(new_adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("New whc")
grid.arrange(plot1, plot2, ncol=2)
```

## Fit with whc in the model

(eq 14) in [Mianabadia, 2020](https://www.sciencedirect.com/science/article/abs/pii/S0022169420305497).

$$
WHC = whc/precipitation\\
AET = (e^{\alpha*WHC*(1-PET^{-1})} -1 )/( e^{\alpha*WHC*(1-PET^{-1})} -  PET^{-1} )
$$

```{r message=FALSE, warning=FALSE}
fit = nls(aet/prec_cond ~ (exp((a*whc/(prec_cond)*(1-(pet/prec_cond)^-1)))-1)/(exp(a*whc/(prec_cond)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = old_adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = old_adf$aet/old_adf$prec_cond, pred= fit_fun(old_adf$pet/old_adf$prec_cond))

plot1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.7, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.65, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("Old whc")
```

```{r message=FALSE, warning=FALSE}
fit = nls(aet/prec_cond ~ (exp((a*whc/(prec_cond)*(1-(pet/prec_cond)^-1)))-1)/(exp(a*whc/(prec_cond)*(1-(pet/prec_cond)^-1))-((pet/prec_cond)^-1)),
          data = new_adf,start = list(a = 2.3))

fit_fun = function(x){
  (exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-x^-1)}

plot_fit = data.frame(obs = new_adf$aet/new_adf$prec_cond, pred= fit_fun(new_adf$pet/new_adf$prec_cond))

plot2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=1,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  geom_text(x= 0.65, y=0.95,
            label = paste0("fitted coef = ",format(coef(fit), digits=3)))+
  ggtitle("New whc")
```

```{r message=FALSE, warning=FALSE}
grid.arrange(plot1, plot2, ncol=2)
```

```{r message=FALSE, warning=FALSE}
fit_fun = function(x){
  (exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*old_adf$whc/(old_adf$prec_cond)*(1-x^-1))-x^-1)}

old_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("Old whc")
```

```{r message=FALSE, warning=FALSE}
fit_fun = function(x){
  (exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-1)/(exp(coef(fit)*new_adf$whc/(new_adf$prec_cond)*(1-x^-1))-x^-1)}

new_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA)  +
  ggtitle("New whc")
```
## Model with vegetation

Starting from the Fu equation, the alpha parameter varying ranging from 2.4 for savannas up to 4 for croplands in new whc and from 3.7 to 9 in old whc.
I create some small data frames in which each type is represented at least 5 times and merge again together.
Each data frames is fitted with his own coefficient.

```{r}
veg <- readRDS(here("data","vegetation.rds"))
new_adf <- left_join(new_adf,veg,by="sitename")
old_adf <- left_join(old_adf,veg,by="sitename")
# vegetation filter (is the same for both)

type <- unique(old_adf[!is.na(old_adf$veg),]$veg)
count <- rep(0,length(type))
for(i in 1:216){
  if(!is.na(old_adf$veg[i])){
      count <- count + (type == old_adf$veg[i])}}

```

```{r}
# old whc filtering

old_adf <- old_adf[old_adf$veg %in% type[count>5],]

old_param <- NULL
for(i in type[count>5]){
  df <- old_adf[old_adf$veg == i,]
  fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = df,start = list(a = 2.5))
    tmp <- data.frame(type = i, coef = coef(fit))
  old_param <- rbind(old_param,tmp)
}

coeff <-c()
for(i in old_adf$veg){
  tmp <- old_param[old_param$type==i,2]
  coeff <- append(coeff,tmp)
}
old_adf$coef <- coeff

fit_fun = function(x,y){
  return( 1 + x - (1 + (x)^y)^(1/y))}

plot_fit = data.frame(obs = old_adf$aet/old_adf$prec_cond, 
                      pred= fit_fun(old_adf$pet/old_adf$prec_cond,old_adf$coef),
                      type= old_adf$veg)

plot1 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.65, y=0.9,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4))) +
  ggtitle("Old whc")
```

```{r}
# new whc filtering

new_adf <- new_adf[new_adf$veg %in% type[count>5],]

new_param <- NULL
for(i in type[count>5]){
  df <- new_adf[new_adf$veg == i,]
  fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = df,start = list(a = 2.5))
    tmp <- data.frame(type = i, coef = coef(fit))
  new_param <- rbind(new_param,tmp)
}

# coefficent updating
coeff <-c()
for(i in new_adf$veg){
  tmp <- new_param[new_param$type==i,2]
  coeff <- append(coeff,tmp)
}
new_adf$coef <- coeff


plot_fit = data.frame(obs = new_adf$aet/new_adf$prec_cond, 
                      pred= fit_fun(new_adf$pet/new_adf$prec_cond,new_adf$coef),
                      type= new_adf$veg)

plot2 <- ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red", se = FALSE)+ 
  geom_text(x= 0.6, y=0.9,
            label = paste0("r^2 = ",format(cor(plot_fit$obs ,plot_fit$pred, use = "complete.obs")^2, digits=4)))+
  ggtitle("New whc")
```

```{r message=FALSE, warning=FALSE}
grid.arrange(plot1, plot2, ncol=2)
```

```{r message=FALSE, warning=FALSE}
old_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond,coef)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("Old whc")
```

```{r message=FALSE, warning=FALSE}
new_adf |> 
 ggplot() +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond, color = "observed")) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_point(aes(x = pet/prec_cond,
                y = fit_fun(pet/prec_cond,coef)
             ,color = "predicted")) +
  ylim(0, NA) +
  xlim(0, NA) +
  ggtitle("New whc")

```

