---
title: "Budyko curve"
output: html_document
date: "2024-02-28"
editor_options: 
  markdown: 
    wrap: sentence
---

## Aim

The aim of this vignette is to create the Budyko curve starting from the data provided on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081)).
To use the vignette properly, download the data and inserted in the folder [data](https://github.com/stineb/rsofundemo/tree/main/data).
The vignette will use the {rsofun} package and will run the p model as the vignette "".
Alternatively, you can load the result of the p model which is stored on [data](https://github.com/stineb/rsofundemo/tree/main/data).

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rsofun)
library(here)
```

```{r}
# use the data found on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081))
driver <- readRDS(here("data","rsofun_driver_data_clean.rds"))
```

```{r}
# load output data
output <- 
```

```{r}
# this chunk is not necessary


# run the model as in the previous vignette
params_modl <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)

# run the model for these parameters
output <- rsofun::runread_pmodel_f(
  driver,
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
output <- output |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len != 1) |>
  select(-len)
```

## Multi year averaging

The Budyko relationship is based on long term measurement.
In this vignette the data will be averaged over the entire time of data collection.

```{r}
# average pet and aet over years 
get_annual_aet_pet <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(aet = sum(aet),
              pet = sum(pet)) |>
    ungroup() |>
    summarise(aet = mean(aet),
              pet = mean(pet))
}

adf <- output |>
  mutate(adf = purrr::map(data, ~get_annual_aet_pet(.))) |>
  unnest(adf) |>
  select(sitename, aet, pet)

# add annual precipitation from forcing
get_annual_prec_cond <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(prec_cond = sum(prec_cond)) |>
    ungroup() |>
    summarise(prec_cond = mean(prec_cond))
}

adf <- driver |>
  unnest(forcing) |>
  left_join(
    output |>
      unnest(data) |>
      select(-snow, -netrad, -fapar, gpp_pmodel = gpp),
    by = c("sitename", "date")
  ) |>
  mutate(prec = (rain + snow) * 60 * 60 * 24) |>
  mutate(prec_cond = prec + cond) |>
  group_by(sitename) |>
  nest() |>
  mutate(adf = purrr::map(data, ~get_annual_prec_cond(.))) |>
  unnest(adf) |>
  select(sitename, prec_cond) |>
  right_join(
    adf,
    by = "sitename"
  )
```

## AET vs precipitation

This figure illustrate the relationship of the AET and precipitation.
Each point represent the multi-year average on a site of AET and precipitation.
Ideally, all the AET value shuold be equal or less than the precipitation value.

```{r}
# AET vs. precipitation plus condensation
adf |>
  ggplot(aes(prec_cond, aet)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  xlim(0, NA) +
  ylim(0, NA)
```

## Curve fitting

The Budyko curve can be modelled following the equation obtain by [FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03) (chinese only).

```math
AET = 1 + PET - (1 + (PET)^(^\omega))(-\omega)
```
To fit the data, a non-linear least square method is used (refer to {nls}).
The only free parameter is \omega.
The fitted equation provides a good estimation of the data, with an r squared of 0.83.

```{r}
# curve fitting and adf updating
#formula: y = 1 + x -( 1+ (x)^a)^1/a 
# ([FU, 1981](http://www.iapjournals.ac.cn/dqkx/en/article/doi/10.3878/j.issn.1006-9895.1981.01.03)) chinese only 

fit = nls(aet/prec_cond ~  1 + pet/prec_cond - (1 + (pet/prec_cond)^a)^(1/a),
          data = adf,start = list(a = 2.3))
fit_fun = function(x){
  return( 1 + x - (1 + (x)^coef(fit))^(1/coef(fit)))}

plot_fit = data.frame(obs = adf$aet/adf$prec_cond, pred= fit_fun(adf$pet/adf$prec_cond))

ggplot(plot_fit, aes(x= obs, y= pred)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_smooth(method = "lm", colour = "red")

```

## Budyko curve

The Budyko curve is shown in this figure along with the fitted model

```{r}
# plotting results

ggplot(adf) +
  geom_point(aes(x = pet/prec_cond, y = aet/prec_cond)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dotted") +
  geom_hline(yintercept = 1, linetype = "dotted") +
  geom_line(aes(x = seq(0.5,4,length.out=dim(adf)[1]),
                y = fit_fun(seq(0.5,4,length.out=dim(adf)[1])))
            ,color = "red") +
  ylim(0, NA) +
  xlim(0, NA)
```
