---
title: "Forcing and validation data"
author: "Francesco Grossi and Beni Stocker"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{P-model usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Aim

This vignette demonstrates the generation of forcing and validation data for {rsofun} (point-scale) simulations.

Two types of simulations are distinguished: 

1. Simulations for existing networks (e.g., FLUXNET) for which forcing data is from local measurements. 
2. Simulations for arbitrary locations for which forcing data is generated from global map data.

Validation data (containing observations that are used to calibrate and test rsofun predictions) may be from different sources, depending on the calibration/testing target variable. Below, the following validation data sources are demonstrated:

- FLUXNET: ecosystem fluxes of water vapour, CO~2~, net radiation (and potentially more, depending on the {rsofun} model scope)
- (more coming soon)

## Simulations for existing networks

### FLUXNET

Variables used as {rsofun} forcing and validation are contained in the same files. Here, we use the FluxDataKit {rsofun} forcing dataset `rsofun_driver_data_clean.rds`, provided on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081)). We also demonstrate data generation using the {FluxDataKit} R package.

#### Forcing data

##### Pre-generated file

We use `rsofun_driver_data_clean.rds`, provided on Zenodo ([Hufkens, 2022](https://doi.org/10.5281/zenodo.8403081)). Download that file and specify its local path. Note: adjust this path depending on where the file is located on your computer.
```{r}
path_forcingfile <- "~/data/FluxDataKit/rsofun_driver_data_clean.rds"
```

This contains data for all available sites, including all meta information required for {rsofun} simulations.
```{r}
df_forcing <- readRDS(path_forcingfile)
```
