In this vignette is possible to run the p_model for multiple sites separately. 
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rsofun)
```
## Data loading

Is possible to run the model with different sites. To do so, refer to [README](https://github.com/stineb/rsofundemo/blob/main/README.md).
The result of the model will be inserted in the folder [fig](https://github.com/stineb/rsofundemo/tree/main/vignettes/fig).

```{r}
# output path for images
out_path <- paste0(getwd(),"/fig/") # path output for the images
sites <- c("ES-Amo","FR-Pue")
drivers <-NULL
for(site in sites){          
  tmp <-  readRDS(paste0("../data/",site,"_p_model_drivers.rda"))
  drivers <- rbind(drivers,tmp)
}
validations <-NULL
for(i in sites){
  tmp <-  readRDS(paste0("../data/",site,"_p_model_validation.rda"))
  validations <- rbind(validations,tmp)
}
```
## Parameter initialization

The initialized parameter will be the same for every site.

```{r}
# define model parameter values from previous
# work
params_modl <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)
```
## Running model

The model is run using `runread_pmodel_f()`. This function takes the nested data structure and runs the model site by site, returning nested model output results matching the input drivers.

```{r}
# run the model for these parameters
output <- NULL
for (i in sites){
  tmp <- rsofun::runread_pmodel_f(
  drivers |> filter(sitename==i),
  par = params_modl)
  output <- rbind(output,tmp)
 } 
```
## Plotting creation

TO show the result, a data frame containing all the sitesis created.

```{r}
# Create data.frame for plotting
df_gpp_plot <- rbind(
  output |>
    unnest(data) |>
    select(date, gpp, sitename) |>
    mutate(type = "P-model output"),
  validations |>
    unnest(data) |>
    select(date, gpp, sitename) |>
    mutate(type = "Observed")
)
df_gpp_plot$type <- factor(df_gpp_plot$type,
                           levels = c('P-model output',
                                      'Observed'))

```

## Output creation

The output is not directly visualized here but is located in [fig](https://github.com/stineb/rsofundemo/tree/main/vignettes/fig). 
The model result is shown separately for each site.

```{r}
# Plot GPP

for(i in sites){
  cairo_pdf(filename = paste0(out_path,i, "_p_model_not_calibrated.pdf"), height=5, width=10)
  png(filename = paste0(out_path,i, "_p_model_not_calibrated.png"), height=700*3, width=700*3, res = 300)
  p = ggplot(data = df_gpp_plot |> filter(sitename==i)) +
  geom_line(
    aes(x = date,
        y = gpp,
        color = type),
    alpha = 0.7
  ) +
    ggtitle(i)+
  scale_color_manual(values = c(
    'P-model output'='grey70',
    'Observed'='black')) +
  theme_classic() +
  theme(panel.grid.major.y = element_line()) +
  labs(
    x = 'Date',
    y = expression(paste("GPP (g C m"^-2, "s"^-1, ")")),
    colour = ""
  )
  p |> print()
  dev.off()
}
```
## Calibration setting

To optimize new parameters based upon driver data and a validation dataset we must first specify an optimization strategy and settings, as well as a cost function and parameter ranges. The strategy is identical for all sites.


```{r}
# calibrating
settings <- list(
  method              = "GenSA",
  metric              = cost_rmse_pmodel,
  control = list(
    maxit = 100),
  par = list(
    kphio = list(lower=0.02, upper=0.2, init = 0.05)
  )
)
```

## Calibrating model parameters

The calibration is done indipendetly for each site.
`rsofun` supports both optimization using the `GenSA` and `BayesianTools` packages. The above statement provides settings for a `GenSA` optimization approach. For this example the maximum number of iterations is kept artificially low. In a real scenario you will have to increase this value orders of magnitude. Keep in mind that optimization routines rely on a cost function, which, depending on its structure influences parameter selection. A limited set of cost functions is provided but the model structure is transparent and custom cost functions can be easily written. More details can be found in the "Parameter calibration and cost functions" vignette.

In addition starting values and ranges are provided for the free parameters in the model. Free parameters include: parameters for the quantum yield efficiency `kphio`, `kphio_par_a` and `kphio_par_b`, soil moisture stress parameters `soilm_thetastar` and `soilm_betao`, and also `beta_unitcostratio`, `rd_to_vcmax`, `tau_acclim` and `kc_jmax` (see `?runread_pmodel_f`). Be mindful that with newer versions of `rsofun` additional parameters might be introduced, so re-check vignettes and function documentation when updating existing code.

With all settings defined the optimization function `calib_sofun()` can be called with driver data and observations specified. Extra arguments for the cost function (like what variable should be used as target to compute the root mean squared error (RMSE) and previous values for the parameters that aren't calibrated, which are needed to run the P-model).



```{r}
# calibrate the model and optimize free parameters

pars <- NULL

for(i in sites){
  tmp <- calib_sofun(
  drivers = drivers |> filter(sitename==i),  
  obs = validations |> filter(sitename==i),
  settings = settings,
  # extra arguments passed to the cost function:
  targets = "gpp",             # define target variable GPP
  par_fixed = params_modl[-1]  # fix non-calibrated parameters to previous 
  # values, removing kphio
  )
  names(tmp) <- paste0(names(tmp),"_",i)
  pars <- append(pars,tmp)
}
```

```{r}
# Update the parameter list with calibrated value


output_new <- NULL
for (i in sites){
  
  params_modl$kphio <- pars[names(pars) == paste0("par_",i)][[1]]
  tmp <- rsofun::runread_pmodel_f(
  drivers|> filter(sitename==i),
  par = params_modl)
  output_new <- rbind(output_new,tmp)
 } 
```

```{r}
# Update data.frame for plotting
df_gpp_plot <- rbind(
  df_gpp_plot,
  output_new |>
    unnest(data) |>
    select(date, gpp, sitename) |>
    mutate(type = "P-model output (calibrated)")
)
df_gpp_plot$type <- factor(df_gpp_plot$type,
                           levels = c('P-model output',
                                      'P-model output (calibrated)',
                                      'Observed'))
```

```{r}
# Plot GPP

for(i in sites){
  cairo_pdf(filename = paste0(out_path,i, "_p_model_calibrated.pdf"), height=5, width=10)
  png(filename = paste0(out_path,i, "_p_model_calibrated.png"), height=700*3, width=700*3, res = 300)
  p = ggplot(data = df_gpp_plot |> filter(sitename==i)) +
  geom_line(
    aes(x = date,
        y = gpp,
        color = type),
    alpha = 0.7
  ) +
  scale_color_manual(values = c(
    'P-model output'='grey70',
    'P-model output (calibrated)'='grey40',
    'Observed'='black')) +
  theme_classic() +
  theme(panel.grid.major.y = element_line()) +
  labs(
    x = 'Date',
    y = expression(paste("GPP (g C m"^-2, "s"^-1, ")")),
    colour = ""
  )
  p |> print()
  dev.off()
}
```
