In this vignettes is possible to run the p_model by selecting one or multiple site.
To work properly be sure to have the "read_meta_fdk.R and" and "generation_validation_bysitename.R"


```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(reshape2)
library(lubridate)
library(FluxDataKit)
library(ncdf4)
library(rsofun)
source("read_meta_fdk.R")
source("generation_validation_bysitename.R")
```

the create_driver_validation is a function into generation_validation_bysitename.R file, 
it will create the data necessary to run the model in the right format.
To work, it needs:
the name acronym of the site
the path file to csv files (the csv file of measurement and the csv file "valid_years_final" must be in the same folder)
the path file to lsm file and nc file
the output path doesn't have to exist since it will be created by the function if not present

```{r}
site <- "FR-Pue"
csv_path <- paste0(getwd(),"/fluxdatakit_oct3/FLUXDATAKIT_FLUXNET/")
lsm_path <- paste0(getwd(),"/fluxdatakit_oct3/FLUXDATAKIT_LSM/")
nc_path <- paste0(getwd(),"/vignettes/ancillary_data/")
out_path <- paste0(getwd(),"/fluxdatakit_oct3/FLUXDATAKIT_RDA/")

# running once per site since the results are saved 
create_driver_validation(site,csv_path,lsm_path,nc_path,out_path)
```

the data can be loaded using readRDS

```{r}
driver <- readRDS(paste0(out_path,site,"_p_model_drivers.rda"))
validation <- readRDS(paste0(out_path,site,"_p_model_validation.rda"))
```


the p model is running in the same way as online

```{r}

# define model parameter values from previous
# work
params_modl <- list(
  kphio              = 0.04998,    # setup ORG in Stocker et al. 2020 GMD
  kphio_par_a        = 0.0,        # set to zero to disable temperature-dependence of kphio
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,  # to recover old setup with soil moisture stress
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,      # value from Atkin et al. 2015 for C3 herbaceous
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)

# run the model for these parameters
output <- rsofun::runread_pmodel_f(
  driver,
  par = params_modl
)
```



```{r}
# Create data.frame for plotting
df_gpp_plot <- rbind(
  output |>
    unnest(data) |>
    select(date, gpp) |>
    mutate(type = "P-model output"),
  validation |>
    unnest(data) |>
    select(date, gpp) |>
    mutate(type = "Observed")
)
df_gpp_plot$type <- factor(df_gpp_plot$type,
                           levels = c('P-model output',
                                      'Observed'))

# Plot GPP
ggplot(data = df_gpp_plot) +
  geom_line(
    aes(x = date,
        y = gpp,
        color = type),
    alpha = 0.7
  ) +
  scale_color_manual(values = c(
    'P-model output'='grey70',
    'Observed'='black')) +
  theme_classic() +
  theme(panel.grid.major.y = element_line()) +
  labs(
    x = 'Date',
    y = expression(paste("GPP (g C m"^-2, "s"^-1, ")")),
    colour = ""
  )
```



```{r}
# calibrating
settings <- list(
  method              = "GenSA",
  metric              = cost_rmse_pmodel,
  control = list(
    maxit = 100),
  par = list(
    kphio = list(lower=0.02, upper=0.2, init = 0.05)
  )
)
```



```{r}
# calibrate the model and optimize free parameters
pars <- calib_sofun(
  drivers = driver,  
  obs = validation,
  settings = settings,
  # extra arguments passed to the cost function:
  targets = "gpp",             # define target variable GPP
  par_fixed = params_modl[-1]  # fix non-calibrated parameters to previous 
  # values, removing kphio
)
```


```{r}
# Update the parameter list with calibrated value
params_modl$kphio <- pars$par["kphio"]

# Run the model for these parameters
output_new <- rsofun::runread_pmodel_f(
  driver,
  par = params_modl
)
```



```{r}
# Update data.frame for plotting
df_gpp_plot <- rbind(
  df_gpp_plot,
  output_new |>
    unnest(data) |>
    select(date, gpp) |>
    mutate(type = "P-model output (calibrated)")
)
df_gpp_plot$type <- factor(df_gpp_plot$type,
                           levels = c('P-model output',
                                      'P-model output (calibrated)',
                                      'Observed'))

# Plot GPP
ggplot(data = df_gpp_plot) +
  geom_line(
    aes(x = date,
        y = gpp,
        color = type),
    alpha = 0.7
  ) +
  scale_color_manual(values = c(
    'P-model output'='grey70',
    'P-model output (calibrated)'='grey40',
    'Observed'='black')) +
  theme_classic() +
  theme(panel.grid.major.y = element_line()) +
  labs(
    x = 'Date',
    y = expression(paste("GPP (g C m"^-2, "s"^-1, ")")),
    colour = ""
  )
```
