---
title: "GPP trend and CO2?"
author: "Beni Stocker"
date: "2024-02-28"
output: 
  html_document:
    code_folding: hide
---

## Data

```{r warning = FALSE, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(rsofun)
library(here)
library(lubridate)
library(FluxDataKit)

if(!require(devtools)){install.packages(devtools)}
devtools::install_github("geco-bern/rsofun")
library(rsofun)
```

Read nice file from FluxDataKit. This contains meteorological data, remotely sensed LAI and fAPAR, and observations-based GPP from an extended set of FLUXNET sites.
```{r warning = FALSE, message = FALSE}
path_forcingfile <- "~/data/FluxDataKit/v3/rsofun_driver_data_v3.rds"
driver <- readRDS(path_forcingfile)
```

Get data from all sites except:

- Croplands
- Wetlands
- Less than 5 years of sequential good-quality data

```{r}
sites <- FluxDataKit::fdk_site_info |>
  dplyr::filter(!(sitename %in% c("MX-Tes", "US-KS3"))) |>  # failed sites
  dplyr::filter(!(igbp_land_use %in% c("CRO", "WET"))) |> 
  left_join(
    fdk_site_fullyearsequence,
    by = "sitename"
  ) |> 
  filter(!drop) |> 
  filter(nyears >= 5)
```

Cut driver data to what's filtered above.
```{r}
driver_forcing_new <- driver |>
  select(sitename, forcing) |> 
  filter(sitename %in% sites$sitename) |> 
  left_join(
    sites |> 
      select(
        sitename, 
        year_start = year_start_fullyearsequence, 
        year_end = year_end_fullyearsequence),
    by = join_by(sitename)
  ) |> 
  unnest(forcing) |> 
  mutate(year = lubridate::year(date)) |> 
  mutate(keep = ifelse(year >= year_start & year <= year_end, TRUE, FALSE)) |> 
  dplyr::select(-year_start, -year_end, -year, -keep) |> 
  group_by(sitename) |> 
  nest()
  
driver <- driver |> 
  filter(sitename %in% sites$sitename)

driver$forcing <- driver_forcing_new$data
```

## Model runs

Run the P-model (rsofun package) assuming 400 ppm for all sites and dates.
```{r warning = FALSE, message = FALSE}
params_modl <- list(
  kphio              = 0.04998,
  kphio_par_a        = 0.0,
  kphio_par_b        = 1.0,
  soilm_thetastar    = 0.6 * 240,
  soilm_betao        = 0.0,
  beta_unitcostratio = 146.0,
  rd_to_vcmax        = 0.014,
  tau_acclim         = 30.0,
  kc_jmax            = 0.41
)

# run the model
output <- rsofun::runread_pmodel_f(
  driver |> 
    mutate(forcing = purrr::map(forcing, ~mutate(., co2 = 400))),
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
output <- output |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len >= 1) |> # at least 3 years data
  select(-len)
```


Model run with varying CO2 as observed.
```{r}
# run the model
output_varco2 <- rsofun::runread_pmodel_f(
  driver,
  par = params_modl
)

# some simulations failed due to missing values in the forcing. If that happens,
# the number of rows in the output data is 1.
output_varco2 <- output_varco2 |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len >= 1) |> # at least 3 years data
  select(-len)
```

This yields data and model outputs for 216 sites. Let's look for trends...

## Trends in observed GPP  {.tabset}

### Annual mean

```{r warning = FALSE, message = FALSE}
get_annual <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(gpp = sum(gpp))
}

# fit linear regression to annual sums for each site and get slope
tmp <- driver |>
  mutate(ndata_gpp = purrr::map_int(forcing, ~sum(!is.na(.$gpp)))) |> 
  filter(ndata_gpp > 0) |> 
  mutate(adf = purrr::map(forcing, ~get_annual(.))) |> 
  mutate(linmod = purrr::map(adf, ~lm(gpp ~ year, data = .))) |> 
  mutate(slope = purrr::map_dbl(linmod, ~coef(.)[2])) |> 
  
  # remove outlier
  filter(slope > -150)
```

Plot distribution of linear trends - no dominance of positive trends.
```{r warning = FALSE, message = FALSE}
tmp |> 
  ggplot(aes(slope, after_stat(count))) +
  geom_histogram(fill = "grey", color = "black", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = median(tmp$slope, na.rm = TRUE), color = "red") +
  theme_classic()
```

### Main growing season

Mean across upper 50% w.r.t. daily GPP values
```{r warning = FALSE, message = FALSE}
get_annual <- function(df){
  threshold <- quantile(df$gpp, probs = 0.5, na.rm = TRUE)
  df |>
    filter(gpp > threshold) |> 
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(gpp = mean(gpp))
}

# fit linear regression to annual sums for each site and get slope
tmp <- driver |>
  mutate(ndata_gpp = purrr::map_int(forcing, ~sum(!is.na(.$gpp)))) |> 
  filter(ndata_gpp > 0) |> 
  mutate(adf = purrr::map(forcing, ~get_annual(.))) |> 
  mutate(linmod = purrr::map(adf, ~lm(gpp ~ year, data = .))) |> 
  mutate(slope = purrr::map_dbl(linmod, ~coef(.)[2]))
```

Plot distribution of linear trends - no dominance of positive trends.
```{r warning = FALSE, message = FALSE}
tmp |> 
  ggplot(aes(slope, after_stat(count))) +
  geom_histogram(fill = "grey", color = "black", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = median(tmp$slope, na.rm = TRUE), color = "red") + 
  theme_classic()
```

### De-seasonalised daily values

TBC

## Trends in LUE  {.tabset}

### Annual mean

```{r warning = FALSE, message = FALSE}
get_annual <- function(df){
  df |>
    mutate(year = lubridate::year(date),
           apar = fapar * ppfd) |> 
    group_by(year) |> 
    summarise(apar = sum(apar),
              gpp = sum(gpp)) |> 
    mutate(lue = gpp/apar)
}

# fit linear regression to annual sums for each site and get slope
tmp <- driver |>
  # filter(sitename != "SE-Svb") |>
  mutate(adf = purrr::map(forcing, ~get_annual(.))) |>
  mutate(linmod = purrr::map(adf, ~try(lm(lue ~ year, data = .)))) |>
  mutate(error = purrr::map_chr(linmod, ~class(.))) |>
  filter(error != "try-error") |>
  mutate(slope = purrr::map_dbl(linmod, ~coef(.)[2])) |>
  filter(slope > -12000) # outlier
```

Plot distribution of linear trends - no dominance of positive trends.
```{r warning = FALSE, message = FALSE}
tmp |>
  ggplot(aes(slope, after_stat(count))) +
  geom_histogram(fill = "grey", color = "black", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = median(tmp$slope), color = "red") + 
  theme_classic()
```

### Main growing season

TBC

### De-seasonalised daily values

TBC

## Trends in P-model  {.tabset}


### Trends in simulated GPP

Combine outputs and forcing.

```{r warning = FALSE, message = FALSE}
tmp <- output_varco2 |>
  select(sitename, data) |>
  unnest(data) |>
  rename(gpp_pmodel = gpp) |>
  left_join(
    driver |>
      select(sitename, forcing) |>
      unnest(forcing),
    by = c("sitename", "date")
  ) |>
  mutate(res = gpp_pmodel - gpp) |>
  group_by(sitename) |>
  nest()
```

Annual mean
```{r warning = FALSE, message = FALSE}
get_annual <- function(df){
  df |>
    mutate(year = lubridate::year(date)) |>
    group_by(year) |>
    summarise(gpp_pmodel = sum(gpp_pmodel))
}

# fit linear regression to annual sums for each site and get slope
tmp <- tmp |>
  mutate(ndata_gpp = purrr::map_int(data, ~sum(!is.na(.$gpp_pmodel)))) |> 
  filter(ndata_gpp > 0) |> 
  mutate(adf = purrr::map(data, ~get_annual(.))) |> 
  mutate(linmod = purrr::map(adf, ~lm(gpp_pmodel ~ year, data = .))) |> 
  mutate(slope = purrr::map_dbl(linmod, ~coef(.)[2]))
```

Plot distribution of linear trends - no clear dominance of positive trends.
```{r warning = FALSE, message = FALSE}
tmp |> 
  ggplot(aes(slope, after_stat(count))) +
  geom_histogram(fill = "grey", color = "black", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = median(tmp$slope, na.rm = TRUE), color = "red") +
  theme_classic()
```

### Trends in residuals

Residuals of model run assuming constant CO2.

Combine outputs and forcing.

```{r warning = FALSE, message = FALSE}
tmp <- output |>
  select(sitename, data) |>
  unnest(data) |>
  rename(gpp_pmodel = gpp) |>
  left_join(
    driver |>
      select(sitename, forcing) |>
      unnest(forcing),
    by = c("sitename", "date")
  ) |>
  mutate(res = gpp_pmodel - gpp) |>
  group_by(sitename) |>
  nest()
```

No need to do annual totals. Fit linear regression directly on residuals.

```{r warning = FALSE, message = FALSE}
# fit linear regression to annual sums for each site and get slope
tmp2 <- tmp |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len >= 1000) |> # at least 3 years data
  select(-len) |>
  mutate(ndata_res = purrr::map_int(data, ~sum(!is.na(.$res)))) |> 
  filter(ndata_res > 0) |> 
  mutate(linmod = purrr::map(data, ~lm(res ~ date, data = .))) |>
  mutate(slope = purrr::map_dbl(linmod, ~coef(.)[2]))
```

Plot distribution of linear trends in residuals - a tendency of negative trends (expected since the model by design doesn't account for the positive CO~2~ effect on GPP).
```{r warning = FALSE, message = FALSE}
tmp2 |>
  ggplot(aes(slope, after_stat(count))) +
  geom_histogram(fill = "grey", color = "black", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = median(tmp2$slope), color = "red") + 
  theme_classic()
```

### Main growing season

```{r warning = FALSE, message = FALSE}
filter_gs <- function(df){
  threshold <- quantile(df$gpp, probs = 0.5, na.rm = TRUE)
  df |>
    mutate(res = ifelse(gpp > threshold, res, NA))
}

# fit linear regression to annual sums for each site and get slope
tmp2 <- tmp |>
  mutate(len = purrr::map_int(data, ~nrow(.))) |>
  filter(len >= 1000) |> # at least 3 years data
  select(-len) |>
  mutate(ndata_res = purrr::map_int(data, ~sum(!is.na(.$res)))) |> 
  filter(ndata_res > 0) |> 
  mutate(data = purrr::map(data, ~filter_gs(.))) |>
  mutate(linmod = purrr::map(data, ~lm(res ~ date, data = .))) |>
  mutate(slope = purrr::map_dbl(linmod, ~coef(.)[2]))
```

Plot distribution of linear trends - no dominance of positive trends.
```{r warning = FALSE, message = FALSE}
tmp2 |>
  ggplot(aes(slope, after_stat(count))) +
  geom_histogram(fill = "grey", color = "black", bins = 30) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_vline(xintercept = median(tmp2$slope), color = "red") + 
  theme_classic()
```

### De-seasonalised daily values

TBC
